---
title: Using R To Automatically Create Dates For Your Syllabus
author: Simon J. Kiss
date: '2018-08-13'
categories: 
    - R
tags: 
    - R
---

Now that August is upon us, academics around the world are sweating a little bit about getting back to the grind of teaching.  

I wanted to introduce a new R package I've developed that does precisely one thing: it creates a text file that contains the days and dates of the classes you'll be teaching. 

Basically I had the idea for this package one late night when I was working on the syllabus for one of the first classes I ever taught. i was proing over Google Calendar, cutting and pasting dates and days, and adding and subtracting numbers and thinking, ``there has to be a way to automate this''. 

The problem was that I was a complete noob not only to R, but to programming in general. Thanks to [this guy](http://politicalscience.uwo.ca/people/faculty/full-time_faculty/dave_armstrong.html) I was able to gin up first a function, then a full package that takes as arguments: the day or days on which your class meets, the first and last days of lectures in your uni's terms. 

Then it prints a text file of those day-date pairs that you can open and plunk into the word processor of your choice and, boom, you can fill in the readings, assignments, etc. as you need.

If you have any comments, criticisms or questions about the package, please write!

The following code illustrates how the code works. Note that it prints the text file to the desktop on a Mac. PC users will have to modify this. 


```{r sample-code, eval=F}
library(devtools)
install_github('sjkiss/syllabus')
createDates(course.days=c('W'), first.day=c('2018/09/01'), last.day=c('2018/12/01'), file='~/Desktop/example-dates-file.txt')
```

And the output looks like this:

```{r print-output, message=F, warning=F, echo=F, results='markup'}
library(devtools)
install_github('sjkiss/syllabus')
library(syllabus)
createDates(course.days=c('W'), first.day=c('2018/09/01'), last.day=c('2018/12/01'))
```

