---
title: "Doug Ford's Win And Sentiment Towards Immigration"
author: 'Simon Kiss'
date: "2018-07-21"
tags: 
   - public-opinion
---

It seems like one of the priorities of the new Ford government is to take a harder line dealing with refugees and asylum-seekers in Canada. Just after taking office, Premier Ford told Prime Minister Trudeau that his new government was basically getting out of the business of supporting the settlement of refugees and blamed the federal government for supposedly encouraging illegal immigration.

This is all a little surprising as the PC campaign did not really make any *explicit* hay out of anti-immigration sentiment. Moreover, as I've pointed out here, Doug Ford's brother was remarkably *popular* among visible minorities as mayor.

The [Laurier Institute for the Study of Public Opinion and Policy](http://lispop.ca) conducted a survey of Ontario voters in the last 10 days of the election campaign and asked a few questions about their views on refugees and immigration.  Here they are broken down by 2018 vote intention. 

```{r setup, echo=F, results='hide', warning=F}
library(knitr)
opts_chunk$set(echo=F, results='hide', fig.width=5, fig.height=4, warning=F, message=F,include=F)
#Set working directory for compilation
opts_knit$set(root.dir='~/OneDrive - Wilfrid Laurier University/projects_folder/ON18/')
```
```{r setwd}
#Set working directory for sourcing
setwd('~/OneDrive - Wilfrid Laurier University/projects_folder/ON18/')
```

```{r load-in}

source('1_load_on18.R')

```

```{r source-economic-immigration-scripts}
names(on18)

source('3_economic_immigration_vote.R')

```
Here are is the distributon of responses for each question on immigration that we asked, broken out by reported vote in the Ontario election. Note, here, I've rescaled each resopnse to range from 0 to 1, with 1 being a pro-immigration response. 
```{r}
library(ggplot2)
theme_set(theme_bw())
```

```{r plot-immigration-vote,include=T}

on18 %>% 
  select(starts_with('imm_'), Vote) %>% 
  select(`Many refugees not real`=imm_refugees, `Immigrants take jobs`=imm_jobs, `Canada keeps criminals out`=imm_criminals, `Overall immigration is netbenefit`=imm_netbenefit, `Vote`=Vote ) %>% 
  gather(.,`Immigration Question` , Response, -Vote) %>% 
  filter(is.na(Vote)==F) %>% 
  ggplot(., aes(x=Vote, y=Response, fill=`Immigration Question`))+geom_boxplot()+theme(axis.text.x=element_text(angle=45, hjust=1))
```
It definitely looks like people who voted for the PCs have concerns about immigration and refugees, more so than voters for other parties. 

One interesting question is whether this varies substantially by region. I've talked to some involved in the election who suggest that this issue was particularly salient in south-western Ontario.  

Moreover, given that we know that *Rob* Ford attracted substantial support from visible minorities in the city of toronto, it would be interesting to see regional variations there. 



To simplify the plot, I took the average of each person's responses to all the immigration questions and filtered out the Greens. 
```{r plot-region-immigration-vote, include=T}


on18 %>% 
  select(imm_rev, Vote, region) %>% 
  filter(is.na(Vote)==F&is.na(region)==F &Vote!='Green') %>% 
  ggplot(., aes(fill=region, y=imm_rev, x=region))+geom_boxplot()+facet_wrap(~Vote)+theme(axis.text.x=element_text(angle=90, hjust=1))

```

These data don't really support the notion that anti-immigration sentiment was concentrated in the southwest, in particular. There was, however, a notable drop amongst PC voters in central and Eastern Ontario, which fits with the notion that rural voters tend to socially conservative views. 

Additionally, PC voters in the 416 (Toronto) did have views on immigration that were close to, although not at,   the average level of the other parties throughout Ontario. 

Lastly, it is worthwhile pointing out that Liberal voters had the most pro-immigration views, the PCs had the least and the NDP straddled the middle. This fits with a general understanding of the NDP as struggling to balance between two types of egalitarians, higher educated, middle class egalitarians, and working class egalitarians with lower levels of education. 

In a future post, I will look at whether the Ontario PCs were able to also capture support from visible minorities (perhaps only in Toronto) in the ways that Rob Ford was able to. 




