---
title: Reporting stepwise regression models with modelsummary
author: Simon Kiss
date: '2023-09-18'
slug: []
categories:
  - R
tags:
  - R
subtitle: ''
summary: ''
authors: []
lastmod: '2023-09-18T10:21:06-04:00'
featured: no
draft: yes
image:
  caption: ''
  focal_point: ''
  preview_only: no
projects: []
---

I love Vincent Arel-Bundock's modelsummary package, and I'm in the process of moving all my model reporting from stargazer to it. I should also say, Vincent is such a super friendly person. Every time I've reached out to him personally with some requests for assistance, he responds. I hope I haven't abused that.

I recently needed to report a series of OLS models for a few different dependent variables. For each dependent variable, I wanted to report a basic model and then two or three more complicated models with more covariates added.  

It wasn't immediately clear how to report this in `modelsummary` but in the end it goes quite quickly. Here's how I did it.

The basic test data I'll use is the dataset on college students' political behaviour contained in Vincent's `Rdatasets` packag 

```{r import-data}
library(tidyverse)
pol<-read_csv(file="https://vincentarelbundock.github.io/Rdatasets/csv/Stat2Data/Political.csv")

```

Let's say I have two dependent variables: did the student vote and did the student feel informed.  For each dependent variable, let's fit two models. A raw one that that checks whether sex was related to either and then whether or not that any relationship holds controlling for a variable measuring students' media consumption habits.


We just make a new vote variable that is 1 for eligible and voted and 0 for eligible and did not vote. 

```{r}
pol %>% 
  mutate(vote2=case_when(
    Vote==4 ~ "Voted",
    Vote==2 ~ "Did not vote",
    Vote==1 ~ "Did not vote",
    TRUE ~ NA_character_
  ))->pol
#Convert to factor
pol$vote2<-factor(pol$vote2, levels=c("Did not vote", "Voted"))

```

Here's my go-to way of doing these kinds of things.

```{r}
#import dataset
pol %>% 
  #Winnow the dataset to variables of interest
  select(Sex, Ethics,Inform, Paper, TV)  %>% 
  # Pivot the dataset so the two dependent variables are in one column
  pivot_longer(cols=c(Ethics, Inform), names_to=c("DV")) %>% 
  #Nest the dataset by each dependent variable
  nest(-DV) %>% 
  #Mutate the dataset addinga list column of models %>% 
  mutate(model1=map(data, function(x) lm(value~Sex, data=x)),
         model2=map(data, function(x) lm(value~Sex+Paper, data=x)), 
         model3=map(data, function(x) lm(value~Sex+TV, data=x)),
         model4=map(data, function(x) lm(value~Sex+TV+Paper, data=x)))->mod.df1
glimpse(mod.df1)
```

So now we have two rows of six columns. Each row contains four models for each separate dependent variable. We could go right to reporting this quite easily.


```{r}
library(modelsummary)

#names(mod.df1$model1)<-c("Ethics", "Informed")
modelsummary(mod.df1$model1, stars=T, 
             gof_omit=c("AIC|BIC|Log.Lik.|RMSE"))
```

And we can report the models with a second set of covariates without too much difficulty. 

Note, it would be nice to figure out how to pipe this immediately to `modelsummary`, but I cannot figure out how to make that work. 

As such, an intermediate step, saving the pivotted models into a new data.frame is necessary.
```{r}
mod.df1 %>% 
  #Pivot the models of interest into one column
  # Store in a new data.frame
  pivot_longer(., cols=c(model1, model2))  ->mod.df1
#Glimpse to see the results
glimpse(mod.df1)

```

This enables easy reporting with `modelsummary`

```{r}
modelsummary(mod.df1$value)
mod.df1
```

Note that models 1 and 2 
