---
title: "Ontario University Entrance Grades, 2005 and 2022"
author: "Simon Kiss"
date: "2026-02-05"
draft: false
execute:
  echo: false
  include: true
  warning: false
  message: false
  fig.path: "images/"
---

```{r}
#| label: setup


library(tidyverse)
library(here)
```

```{r}
#| label: import-data


df1<-read.csv(file=here("posts/2026-02-05-ontario-entrance-averages/data/fall_2005.csv"))
df2<-read.csv(file=here("posts/2026-02-05-ontario-entrance-averages/data/fall_2022.csv"))

```

There is some discussion today of grade inflation in the province of Ontario ([link](https://www.cbc.ca/news/canada/toronto/ontario-high-school-grades-on-the-rise-9.7072730)). It seems like it's a real thing to me but I have always wanted to poke around a bit to get some leverage on it. Ontario universities offer the [Common University Data Ontario](https://ontariosuniversities.ca/open-data/common-university-data-ontario-cudo/) database that publishes a lot of information about Ontario universities, including entrance averages by field of study and by university.

I took the data from 2005 and 2022, which was the last year available and graphed it in a few ways.

```{r}
#| label: inspect


df1 %>% 
  rename("University"=1, "Agriculture"=2, "Architecutre"=3,"Arts"=4, "Business"=5, "Education"=6, "Engineering"=7, "Environmental Studies"=8, "Fine Arts"=9, "Household Science"=10, "Journalism"=11, "Landscape Architecture"=12, "Mathematics"=13, 
"Music"=14, "Nursing"=15, "Administration"=16, "Physical Education"=17, "Rehab Med"=18, "Science"=19, "Social Work"=20, "Other"=21, "Overall"=22)->df1

df2 %>% 
  rename("University"=1, "Agriculture"=2, "Architecutre"=3, "Biological/Biomedical Sciences"=4,"Business"=5,"Journalism"=6,"Computer Science"=7,"Education"=8, "Engineering"=9, "Household Science"=10, "Fine Arts"=11, "Science"=12, "Health Professionals"=13, "Physical Education"=14, "Languages / Linguistics"=15, "Arts"=16, "Mathematics"=17, "Music"=18, "Natural Resources / Conservation"=19, "Nursing"=20, "Physical Science"=21, "Psychology"=22,"Social Sciences"=23,"Other"=24, "N/A"=26, "Overall"=27,"Diplomas"=25) ->df2

```

```{r}
#| label: data-orgnization

#pivot each data set so each field of study is in one column
df1 %>% pivot_longer(., cols=-University) %>% 
  #add a year
  mutate(Year=rep(2005, nrow(.)))->df1
df2 %>% pivot_longer(., cols=-University) %>% 
  mutate(Year=rep(2022, nrow(.))) %>% 
  #bind rows together into dataset on
  bind_rows(df1)->on
#Now repivot wider the full data set one column for 2005 and one column for 2022
on %>% 
  pivot_wider(id_cols=c(University,name), names_from=c("Year"), values_from=c("value")) %>% 
  #We can only use cases where we have equal names
  filter(complete.cases(.)) ->on
```

Here is the graph of the overall averages by university. This is the best comparison, probably.
Western University and University of Toronto had the highest averages in both years and those grades went up by about eight percentage points, just eyeballing.  Interestingly, York University's entrance average stayed pretty flat. 
```{r}
#| label: overall
on %>%    
  filter(str_detect(name, "Overall")) %>%
  ggplot(., aes(y=`2005`, yend=`2022`,x=1.3, xend=2.7))+geom_segment(col="lightgrey")+labs(y="Grade", x="Year")+theme_minimal()+theme(legend.position="bottom", panel.grid = element_blank())+
  scale_x_continuous(breaks=c(1.3, 2.7), labels=c("2005", "2022"), limits=c(1,3))+geom_point(
    aes(y=`2005`,x=1.3,  col=University)
  )+
  geom_point(
     aes(y=`2022`,x=2.7,  col=University)
  )


```

Here is a more elaborate graph that shows grade inflation only by field of study, averaged by university. The only field that seems to be holding the line is mathematics. 

```{r}
#| label: field-of-study
on %>%    
  filter(str_detect(name, "Overall", negate=T)) %>%
  group_by(name) %>% 
  mutate(`2005`=mean(`2005`),
         `2022`=mean(`2022`)) %>% 
  select(-University) %>% 
  distinct() %>% 
  ggplot(., aes(y=`2005`, yend=`2022`,x=1.3, xend=2.7))+geom_segment(col="lightgrey")+labs(y="Grade", x="Year")+theme_minimal()+theme(legend.position="bottom", panel.grid = element_blank())+
  scale_x_continuous(breaks=c(1.3, 2.7), labels=c("2005", "2022"), limits=c(1,3))+geom_point(
    aes(y=`2005`,x=1.3,  col=name)
  )+
  geom_point(
     aes(y=`2022`,x=2.7,  col=name)
  )
```

And here is the full plot: field of study by university, excluding agriculture and household sciences because there are so few points.

```{r}
#| label: draw-plot

on %>%    
  filter(str_detect(name, "Agriculture|Household Science", negate=T)) %>% 
  ggplot(., aes(y=`2005`, yend=`2022`,x=1.3, xend=2.7,  col=name))+geom_segment()+facet_wrap(~University)+labs(y="Grade", x="Year")+theme_minimal()+theme(legend.position="bottom", panel.grid = element_blank())+
  scale_x_continuous(breaks=c(1.3, 2.7), labels=c("2005", "2022"), limits=c(1,3))+geom_point(
    aes(y=`2005`,x=1.3,  col=name)
  )+
  geom_point(
     aes(y=`2022`,x=2.7,  col=name)
  )

```

